A SAT-Based Counterexample Guided Method for Unbounded Synthesis

### 2016-01-26 paper.tex
Edits based on Leonid's feedback

\documentclass{llncs}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{fixltx2e}
\usepackage{subcaption}
\captionsetup{compatibility=false}
\usepackage{tikz}
\usetikzlibrary{trees}

@@

The bounded synthesis algorithm begins with an empty game as seen in Figure
\ref{fig:emptytree}.  Initially we are playing on behalf of the environment
because it chooses the first move in each step.  The empty game is passed to
the SAT solver, which searches for a candidate environment strategy.  If a
candidate is found (Figure \ref{fig:candidate}) then it is checked for a
spoiling strategy by solving for the controller.  If no spoiling strategy
exists, that means our candidate is a winning strategy and the algorithm
terminates.  Otherwise we find a counterexample (Figure \ref{fig:cex}), which
is used to refine the empty game tree to include the first move from the
controller's spoiling strategy (Figure \ref{fig:refined}).  The algorithm
continues by finding a new candidate for the environment (Figure
\ref{fig:candidate2}), and a new counterexample to refine the game abstraction
once again (Figure \ref{fig:refined2}). The full procedure is listed in
Algorithm \ref{alg:bounded}.

@@

    \begin{algorithmic}
        \algrestore{b1}
        \Function{findCandidate}{$p, \langle s, k \rangle, T$}
            \State $f \gets $ \IfElse{$p = \texttt{cont}$}{\Call{treeFormula}{$T$}}{\Call{\textoverline{treeFormula}}{$T$}} \EndIfElse
            \State $sol \gets $ \Call{SAT}{$s(X_T) \land f$}
            \IIf{$sol = \texttt{unsat}$} 
                \Return $\emptyset$ \Comment{No candidate exists}
            \EndIIf
            \State \Return $\{ \langle n, m \rangle | n \in $ \Call{gtNodes}{T} $, m = \Call{sol}{n} \}$ \Comment{Fix candidate moves in $T$}
        \EndFunction
        \algstore{b2}
    \end{algorithmic}

@@

@@

\begin{algorithm}
    \begin{algorithmic}[1]
        \Function{solveUnbounded}{$T$}
            \State $B^M \gets E$
            \State $B^m[0] \gets E$
            \For{$k = 1 \dots$}
                \IIf{\Call{SAT}{$I \land B^M$}}
                    \Return \texttt{unrealisable} \Comment{Losing in the initial state}
                \EndIIf
                \IIf{$\exists i \  B^m[i] \equiv B^m[i+1]$}
                    \Return \texttt{realisable} \Comment{Reached fixed point}
                \EndIIf
                \State $B^m[k] \gets \top$
                \State \Call{checkRank}{$I, k$}
            \EndFor
        \EndFunction
        \algstore{u1}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{u1}
        \Require \emph{May} and \emph{must} invariants hold
        \Ensure \emph{May} and \emph{must} invariants hold
        \Ensure $\lnot s \in B^m[k]$ if there is a winning controller strategy of length $k$ starting at $s$
        \Ensure $s \in B^M$ if there is a winning environment strategy of length $k$ starting at $s$
%%%        $\exists u_{k..1} \forall c_{k..1} \  s(x_k) \land (E_k \lor (\delta(x_k, u_k, c_k) \land E(x_{k-1}) \lor ... \land E(x_1) \lor (\delta(x_1, u_1, c_1) \land E(x_0))...)$
%%%            $\forall u_{k..1} \exists c_{k..1} \  s(x_k) \land \lnot E_k \land \delta(x_k, u_k, c_k) \land \lnot E(x_{k-1}) \land ... \land \delta(x_1, u_1, c_1) \land \lnot E(x_0)$
        \Function{checkRank}{$s, k$}
            \State \Return \Call{solveAbstract}{$\texttt{env}, \langle s, k \rangle, \emptyset$}
        \EndFunction
        \algstore{u2}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{u2}
        \Function{solveAbstract}{$p, \langle s, k \rangle , T$}
        \State $cand \gets $ \Call{findCandidate}{$p, \langle s, k \rangle, T$} \Comment{Look for a candidate}
        \IIf{$k = n - 1$} \Return $cand$ \EndIIf \Comment{Reached the bound}
        \State $T' \gets T$
        \Loop
            \IIf{$cand = \emptyset $} \Return $\emptyset $ \EndIIf \Comment{No candidate: return with no solution}
            \State $\langle cex, l, u \rangle \gets $ \Call{verify}{$p, \langle s, k \rangle, T, cand$} \Comment{Verify candidate}
            \IIf{$cex = \False$} \Return $cand$ \EndIIf \Comment{No counterexample: return candidate}
            \State $T' \gets $ \Call{gtAppend}{$T', l, u$} \Comment{Refine AGT with counterexample}
            \State $cand \gets $ \Call{solveAbstract}{$p, \langle s, k \rangle, T'$} \Comment{Solve refined AGT}
        \EndLoop
        \EndFunction
        \algstore{u3}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{u3}
        \Function{findCandidate}{$p, \langle s, k \rangle, T$}
            \State $f \gets $ \IfElse{$p = \texttt{cont}$}{\Call{treeFormula}{$T$}}{\Call{\textoverline{treeFormula}}{$T$}} \EndIfElse
            \State $sol \gets $ \Call{SAT}{$s(X_T) \land f$}
            \If{$sol = \texttt{unsat}$} 
                \State $\sigma \gets $ \Call{generalise}{$s$} \Comment{Expand $s$ to a set of states}
                \State \Call{learn}{$\sigma, T$}
                \State \IfElse{$p = \texttt{cont}$}{\Call{learn}{$\sigma, T$}}{\Call{\textoverline{learn}}{$\sigma, T$}} \EndIfElse
                \State \Return $\emptyset$ \Comment{No candidate exists}
            \Else
                \State \Return $\{ \langle n, c \rangle | n \in $ \Call{gtNodes}{T} $, c = \Call{sol}{n} \}$ \Comment{Fix candidate moves in $T$}
            \EndIf
        \EndFunction
        \algstore{u4}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{u4}
        \Function{verify}{$p, \langle s, k\rangle, gt, cand$}
            \For{$l \in leaves(gt)$}
            \State $\langle k', s'\rangle \gets $ \Call{outcome}{$\langle s, k\rangle, gt, l$} \Comment{Get rank and state at leaf}
                \State $u \gets $ \Call{solveAbstract}{\Call{opponent}{$p$}, $\langle k', s' \rangle, \emptyset$} \Comment{Solve for the opponent}
                \IIf{$u \neq \emptyset$} \Return $\langle \False, l, u \rangle$ \EndIIf \Comment{Return counterexample}
            \EndFor
            \State \Return $\langle \True, \emptyset, \emptyset \rangle$
        \EndFunction
    \end{algorithmic}

    \caption{Unbounded Synthesis}
    \label{alg:unbounded}
\end{algorithm}

@@

The unbounded synthesis algorithm (Algorithm \ref{alg:unbounded}) effectively
consists of two communicating solvers: the counterexample-guided bounded
reachability solver and the unbounded solver based on incremental induction.
Fortunately, correctness of both solvers can be established largely
independently, which simplifies the proof of correctness of the overall
algorithm.

@@

>>>>

\documentclass{llncs}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{fixltx2e}
\usepackage{subcaption}
\captionsetup{compatibility=false}
\usepackage{tikz}
\usetikzlibrary{trees}
\usetikzlibrary{positioning}

@@

The bounded synthesis algorithm (Algorithm \ref{alg:bounded}) begins with an
empty game as seen in Figure \ref{fig:emptytree}.  Initially we are playing on
behalf of the environment because it chooses the first move in each step.  The
empty game is passed to the SAT solver, which searches for a candidate
environment strategy.  If a candidate is found (Figure \ref{fig:candidate})
then it is checked for a spoiling strategy by solving for the controller.  If
no spoiling strategy exists, that means our candidate is a winning strategy and
the algorithm terminates.  Otherwise we find a counterexample (Figure
\ref{fig:cex}), which is used to refine the empty game tree to include the
first move from the controller's spoiling strategy (Figure \ref{fig:refined}).
The algorithm continues by finding a new candidate for the environment (Figure
\ref{fig:candidate2}), and a new counterexample to refine the game abstraction
once again (Figure \ref{fig:refined2}). The full procedure is listed in
Algorithm \ref{alg:bounded}.

@@

    \begin{algorithmic}
        \algrestore{b1}
        \Function{findCandidate}{$p, \langle s, k \rangle, T$}
            \State $f \gets $ \IfElse{$p = \texttt{cont}$}{\Call{treeFormula}{$T$}}{\Call{\textoverline{treeFormula}}{$T$}} \EndIfElse
            \State $sol \gets $ \Call{SAT}{$s(X_T) \land f$}
            \If{$sol = \texttt{unsat}$} 
                \If{\texttt{unbounded}} \Comment{Active only in the unbounded solver}
                    \State $\sigma \gets $ \Call{generalise}{$s$} \Comment{Expand $s$ to a set of states}
                    \State \IfElse{$p = \texttt{cont}$}{\Call{learn}{$\sigma, T$}}{\Call{\textoverline{learn}}{$\sigma, T$}} \EndIfElse
                \EndIf
                \State \Return $\emptyset$ \Comment{No candidate exists}
            \Else
                \State \Return $\{ \langle n, c \rangle | n \in $ \Call{gtNodes}{T} $, c = \Call{sol}{n} \}$ \Comment{Fix candidate moves in $T$}
            \EndIf
        \EndFunction
        \algstore{b2}
    \end{algorithmic}

@@

\subsection{Extending Bounded Synthesis to Unbounded Games}

\begin{algorithm}[h]
    \begin{algorithmic}[1]
        \Function{solveUnbounded}{$T$}
            \State $B^M \gets E$
            \State $B^m[0] \gets E$
            \For{$k = 1 \dots$}
                \IIf{\Call{SAT}{$I \land B^M$}}
                    \Return \texttt{unrealisable} \Comment{Losing in the initial state}
                \EndIIf
                \IIf{$\exists i \  B^m[i] \equiv B^m[i+1]$}
                    \Return \texttt{realisable} \Comment{Reached fixed point}
                \EndIIf
                \State $B^m[k] \gets \top$
                \State \Call{checkRank}{$I, k$}
            \EndFor
        \EndFunction
        \algstore{u1}
    \end{algorithmic}

    \begin{algorithmic}
        \algrestore{u1}
        \Require \emph{May} and \emph{must} invariants hold
        \Ensure \emph{May} and \emph{must} invariants hold
        \Ensure $\lnot s \in B^m[k]$ if there is a winning controller strategy of length $k$ starting at $s$
        \Ensure $s \in B^M$ if there is a winning environment strategy of length $k$ starting at $s$
%%%        $\exists u_{k..1} \forall c_{k..1} \  s(x_k) \land (E_k \lor (\delta(x_k, u_k, c_k) \land E(x_{k-1}) \lor ... \land E(x_1) \lor (\delta(x_1, u_1, c_1) \land E(x_0))...)$
%%%            $\forall u_{k..1} \exists c_{k..1} \  s(x_k) \land \lnot E_k \land \delta(x_k, u_k, c_k) \land \lnot E(x_{k-1}) \land ... \land \delta(x_1, u_1, c_1) \land \lnot E(x_0)$
        \Function{checkRank}{$s, k$}
            \State \Return \Call{solveAbstract}{$\texttt{env}, \langle s, k \rangle, \emptyset$}
        \EndFunction
    \end{algorithmic}
    \caption{Unbounded Synthesis}
    \label{alg:unbounded}
\end{algorithm}

The unbounded synthesis algorithm (Algorithm \ref{alg:unbounded}) effectively
consists of two communicating solvers: the counterexample-guided bounded
reachability solver and the unbounded solver based on incremental induction.
The unbounded solver calls the bounded solver with an increasing bound (line 8)
while learning states from abstract game trees that either must or may lose for
the controller. When the initial state is contained within the must-losing
states the game is known to be unrealisable (line 5).

May-losing states are learned by removing controller-winning states. We
maintain a set of states for every rank up the current bound. The states we
learn while executing the bounded algorithm can only said to be
controller-winning for a particular rank or below. However, we construct each
set carefully such that it is monotonically larger than the set of the rank
below (see Figure \ref{fig:maylosing}) and so that the environment is unable to
force play from one set to the next. Due to this careful construction, when two
adjacent sets become equivalent we know that the algorithm has reached a fixed
point and the controller is winning in the unbounded game (line 6).

\begin{figure}
    \centering
    \begin{tikzpicture}
        \node [minimum size=2cm,circle,draw] (a){$B^m[n]$};
        \node [minimum size=1.5cm,circle,draw,right=1.8cm of a] (b){$B^m[n-1]$};
        \node [minimum size=1cm,right=1.5cm of b] (dots){$\dots$};
        \node [minimum size=0.5cm,circle,draw,right=1.5cm of dots] (c){$B^m[0]$};
        \draw[-latex] (a.east) -- node [above] {$\supset$} (b.west);
        \draw[-latex] (b.east) -- node [above] {$\supset$} (dots);
        \draw[-latex] (dots.east) -- node [above] {$\supset$} (c.west);
    \end{tikzpicture}
    \caption{May-losing states}
    \label{fig:maylosing}
\end{figure}

When executing the unbounded solver, lines 18 and 19 become active in the
bounded solver. These lines call the learning procedures when the solver fails
to find a candidate for an abstract game tree. The states symbolically
represented by nodes in the tree are losing for whichever player could not find
a winning candidate and can be extracted from the tree using interpolation.

@@

@@

The correctness of the unbounded synthesis algorithm can be established
independently from that of the bounded algorithm. Fortunately, the correctness
of the bounded solver has been established in \cite{narodytska2014}.

@@

We evaluate our approach on the benchmarks of the 2015 synthesis competition
(SYNTCOMP'15).
